Demography and linked selection in wild and domesticated maize

### 2015-08-20 manuscript.tex
updates

\begin{abstract}
  Unique selective and demographic operate on domesticated plant species. These forces interact during and after domestication to generate the patterns of DNA variability that are persistent in crop species today. To quantify the interplay between demography and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. We utilized whole genome sequence data from 23 maize and 13 teosinte individuals to make inferences.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps, specifically positive selection on new genic mutations, are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels superior to those seen in teosinte. Our results demonstrate that in domesticated species, demographic and selective history in the ancient and recent past both contribute to genetic variability that is present today, providing substantial implications for the continued improvement of domesticated species.

@@

Our initial comparison of synonymous and missense substitutions has low power to detect the effects of selection acting on multiple mutations or standing genetic variation, because in such cases diversity is not necessarily reduced \cite{innan2004,messer2013}. Such ``soft sweeps'', however, are still expected to occur more frequently in functional regions of the genome and could provide an alternative explanation for the observed reduction of diversity in genes.
To test this possibility, we performed a genome-wide scan for selection using a method expected to be reasonably sensitive to both hard and soft sweeps \cite{garud2015}. 
Even after removing genes in the top 20\% of the H12 statistic used to identify targets of selection, however, qualitative differences between maize and teosinte in patterns of diversity in and near genes remained unchanged (Figure \ref{sFig:H12}).

@@

Finally, we reconstructed the demographic history of maize and teosinte using a recently developed coalescent approach \citex that does not require conditioning on a particular model (Figure \X).
Though this analysis does highlight a potential weakness with our assumption of a constant size for teosinte, it nonetheless is broadly consistent with our simple model,   identifying a clear domestication bottleneck followed by rapid population expansion (Figure \X).

\subsection{Population expansion leads to stronger purifying selection in modern maize}
Motivated by the rapid post-domestication expansion of maize evident from our demographic analysis, we investigated whether patterns of diversity at low frequency --- and therefore on average younger --- polymorphisms would show similar results in comparisons of maize and teosinte.
We thus repeated our analysis of diversity using alleles present in only a single individual in the sample.
In direct contrast to patterns observed for pairwise nucleotide diversity (Figure \ref{fig:purify}A), diversity of singletons was more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. %\jri{we should test singleton curve for genes not in 20\% highest H12 also!} I don
Finally, while direct comparison of curves for pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

Another possibility is that maize and teosinte don't show evidence of abundant hard sweeps because instead positive selection tends to operate in the form of soft sweeps. The theory of when soft sweeps are expected was worked out by \cite{messer2013population}. They showed that for $\theta_b = 2N\mu_b$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b > 0$, if $\theta_b > \frac{1}{log({Ns_b})}$, then soft sweeps in the form of adaptation from multiple mutations should be common. Conversely, for neutral sites that somehow become beneficial (e.g. change in environment), if $\theta_b > \frac{1}{log({2Ns_b})}$ then soft sweeps in the form of adaptation from standing neutral variation should be common. Alternatively, for previously deleterious sites that become beneficial, if $\theta > \frac{1}{log \left(1+\frac{2Ns_b}{2Ns_d + 1} \right)}$, where $s_d$ is the previous selection coefficient, soft sweeps again the form of adaptation from standing variation should be common. The implications of this theory are that soft sweeps are common when $\theta_b \geq 1$, which can be achieved through a large population size or a high mutation rate. If $\mu = 3 \times 10^{-8}$  \cite{clark2005}, this implies a population size of $\geq$ 16.67 million, which our demographic analysis shows is larger than that of teosinte but not out of the question for modern maize.

@@

\subsection{Demography influences the efficiency of purifying selection}
The observation that maize pairwise diversity is less impacted by the distance to the nearest gene than is teosinte pairwise diversity is reasonable from a long-term evolutionary standpoint. Theory has established that purifying selection is more efficient in a large population than in a small one \cite{kimura1984}, and this observation most likely reflects that prediction. More specifically, if teosinte $N_e$ remained relatively constant while maize bottlenecked and recovered exponentially, this provides that the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte, regardless of how much maize has ballooned in the recent past. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte. \jri{shorten this!}

The reversal of this trend when we analyze only singleton diversity instead of pairwise diversity, however, stands out as a notable observation. Every mutation begins as a singleton (an allele present in only one individual), and therefore singletons are, on average, the youngest class of alleles that can be observed. Therefore unlike pairwise patterns of diversity, which are most heavily influenced by intermediate frequency alleles based on the definition of $\pi$ \cite{nei1979}, singleton diversity is most influenced by recent patterns of evolution. Hence, because our demographic estimation indicates dramatic expansion of maize $N_e$ in the recent past, we expect for purifying selection to presently operate more efficiently in maize than in teosinte. This observation is very clearly demonstrated by the fact that singleton diversity in maize is more impacted by the distance to the nearest gene than it is in teosinte, as was shown in Figure \ref{fig:purify}. \jri{shorten!}

A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. \jri{this only works for weakly selected mutations but is an interesting point. need to also point out that this means new mutations in maize likely much more strongly selected than new mutations in teosinte, perhaps with implications for breeding?}

@@

Finally, \jri{MSMC text here}

@@

\bibliography{Reference}
\bibliographystyle{pnas}
\onecolumn
\input{supportingInfo} 
%\nocite{*}

\end{article}
\end{document}

>>>>

\begin{abstract}
  Unique selective and demographic forces operate on domesticated plant species. These forces interact during and after domestication to generate the patterns of DNA variability that are persistent today. To quantify the interplay between demography and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. Our sample included whole genome sequence data from 23 maize and 13 teosinte individuals.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection at purging deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels superior to those seen in teosinte. Our results demonstrate that particularly in domesticated species or bottlenecked species, demographic and selective history in the ancient and recent past both contribute to genetic variability that is present today, providing substantial implications for the continued improvement of domesticated species.

@@

Our initial comparison of synonymous and missense substitutions has low power to detect the effects of selection acting on multiple mutations or standing genetic variation, because in such cases diversity is not necessarily reduced \cite{innan2004,messer2013}. Such ``soft sweeps'', however, are still expected to occur more frequently in functional regions of the genome and could provide an alternative explanation for the observed reduction of diversity in genes. To test this possibility, we performed a genome-wide scan for selection using a method expected to be reasonably sensitive to both hard and soft sweeps \cite{garud2015}; 
after removing genes in the top 20\% of the H12 statistic used to identify targets of selection, we observe that qualitative differences between maize and teosinte in patterns of diversity in and near genes remained unchanged (Figure \ref{sFig:H12}A).

@@

\color{blue}
Finally, we reconstructed the demographic history of maize and teosinte using a recently developed coalescent approach \citex that does not require conditioning on a particular model (Figure \X).
Though this analysis does highlight a potential weakness with our assumption of a constant size for teosinte, it nonetheless is broadly consistent with our simple model,   identifying a clear domestication bottleneck followed by rapid population expansion (Figure \X). 
\color{black}

\subsection{Population expansion leads to stronger purifying selection in modern maize}
Motivated by the rapid post-domestication expansion of maize evident from our demographic analysis, we investigated whether patterns of diversity at low frequency --- and therefore on average younger --- polymorphisms would show similar results in comparisons of maize and teosinte.
We thus repeated our analysis of diversity using alleles present in only a single individual in the sample.
In direct contrast to patterns observed for pairwise nucleotide diversity (Figure \ref{fig:purify}A), diversity of singletons was more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This relationship was maintained when we analyzed diversity surrounding genes that did not show evidence of hard or soft sweeps according to H12 (Figure \ref{sFig:H12})
Finally, while direct comparison of curves for pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

Another possibility is that maize and teosinte don't show evidence of abundant hard sweeps because instead positive selection tends to operate in the form of soft sweeps. The theory of when soft sweeps are expected was worked out by \cite{messer2013population}. They showed that for $\theta_b = 2N\mu_b$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b > 0$, if $\theta_b > \frac{1}{log({Ns_b})}$, then soft sweeps in the form of adaptation from multiple mutations should be common. Conversely, for neutral sites that somehow become beneficial (e.g. change in environment), if $\theta_b > \frac{1}{log({2Ns_b})}$ then soft sweeps in the form of adaptation from standing neutral variation should be common. Alternatively, for previously deleterious sites that become beneficial, if $\theta > \frac{1}{log \left(1+\frac{2Ns_b}{2Ns_d + 1} \right)}$, where $s_d$ is the previous selection coefficient, soft sweeps again the form of adaptation from standing variation should be common. The implications of this theory are that soft sweeps are common when $\theta_b \geq 1$ \cite{messer2013population}, which can be achieved through a large population size or a high mutation rate. The best estimate of mutation rate for maize is $\mu = 3 \times 10^{-8}$  \cite{clark2005}, which implies that soft sweeps should be common if population size is $\geq$ 16.67 million. According to our demographic analysis, this is larger than that of teosinte but not out of the question for modern maize.

@@

\subsection{Demography influences the efficiency of purifying selection}
The observation that maize pairwise diversity is less impacted by the distance to the nearest gene than is teosinte pairwise diversity is reasonable from a long-term evolutionary standpoint. Purifying selection works more efficiently in a large population than in a small one \cite{kimura1984}, so this observation likely reflects that prediction. Since maize bottlenecked before recovering exponentially, the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte.

The reversal of this trend when we analyze only singleton diversity instead of pairwise diversit stands out as a notable. Every mutation begins as a singleton (an allele present in only one individual), and therefore singletons are, on average, the youngest class of alleles that can be observed. Unlike pairwise patterns of diversity, which are most heavily influenced by intermediate frequency alleles based on the definition of $\pi$ \cite{nei1979}, singleton diversity is most influenced by recent patterns of evolution. Hence, because our demographic estimation indicates dramatic expansion of maize $N_e$ in the recent past, we expect for purifying selection to presently operate more efficiently in maize than in teosinte. This observation is very clearly demonstrated by the fact that singleton diversity in maize is more impacted by the distance to the nearest gene than it is in teosinte, as was shown in Figure \ref{fig:purify}.

A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. \textcolor{red}{breeding implications?}

@@

\color{blue}
Finally, MSMC text here
\color{black}

@@

\bibliography{Reference}
\bibliographystyle{pnas}
\onecolumn
\input{supportingInfo}

\end{article}

\end{document}
